Prediction of Protein-Protein Interactions on the Human and Rice Interactome

### 2020-08-14 camila/Report/article_CR.tex

\begin{abstract}
A gene co-expression network is an undirected graph, where each node corresponds to a gene. A pair of nodes is connected if they show similar expression patterns. These co-expression networks are of biological interest since the co-expressed genes are usually controlled by the same transcriptional regulatory pathway, functionally related or members of the same pathway or metabolic complex. Weighted gene co-expression network analysis (WGCNA)  is a method of data mining widely used to build co-expression networks, identify groups (modules) of highly correlated genes and relate these modules to external traits.

The aim of this work is to identify rice genes that respond to saline stress. The methodology is based on WGCNA with a new approach in the modules detection, using  the hierarchical link clustering (HLC) technique that allows the recognition of overlapping communities, which may have more biological meaning given the overlapping regulatory domains of systems that generate co-expression \cite{gaiteri2014beyond}. Finally the LASSO method is used to select the most significant modules associated with rice phenotypical responses to salt stress. The genes differentially expressed within the selected modules are enriched with gene ontology annotations and their interaction networks are studied.
%\keywords{First keyword  \and Second keyword \and Another keyword.}
% need to focus on the main results, need to say what your main findings are - the set of genes, and how they are validated
\end{abstract}

\section*{Introduction}
Introduction text

\section{Methodology}
\subsection{Data pre-processing} 
%-----------------------------
% RNA-seq data
%-----------------------------
RNA-seq data was accessed through GEO database \cite{GEOAcces90:online} (Accession number GSE98455), corresponding to $n=57845$ gene expression profiles of shoot tissues measured for both control and salt condition in $p=92$ accessions of the Rice Diversity Panel 1 (with two biological repetitions). \\

The RNA-seq data cannot be directly interpreted, therefore a normalization process has to be done to deal with the various biases that affect quantification results. The normalization technique used was DESeq2 \cite{love2014moderated}. From the normalized data, the two biological repetitions of each accession were averaged and genes exhibiting low variance (the ratio of upper quantile to lower quantile smaller than 1.5) or low expression (more than 80\% samples with values smaller than 10) were removed. At this point, control and salinity stress treatment data are separated into two matrices $C=[c_{ij}]_{n \times p}$ and $T=[t_{ij}]_{n \times p}$, respectively, where $c_{ij}$ and $t_{ij}$ represent normalized expression level of the gene $i$ in the accession $j$.

Changes in gene expression between control and stress conditions, are measured in terms of log ratios. Matrix $E=[e_{ij}]_{n \times p}$, known as the Log Fold Change matrix, is computed by setting $e_{ij}=\log_2 (t_{ij}/c_{ij})$.

@@

\subsection{Network construction}
The co-expression network is constructed based on WGCNA, as explained below.\\

The Log Fold Change matrix is used to construct the weighted co-expression network using the absolute value of the Pearson correlation as the similarity measure between genes. The matrix $S=[s_{ij}]_{n\times n}$ measures the level of concordance between gene expression profiles across experiments and is transformed into an adjacency matrix $A=[a_{ij}]_{n\times n}$ where $a_{ij} = (s_{ij})^\beta $ encodes the connection strength between each pair of genes. In other words, the elements of the adjacency matrix are the similarity values up to the power $\beta > 1$ so the degree distribution will fit a scale-free network. This kind of networks contain many nodes with very few connections and a small number of hubs with high connections. \\

@@

\subsection{Module detection}	
Once the network has been constructed, the next step is to study its structure and dynamics identifying communities also called modules. Membership in these modules may overlap in biological contexts, where modules may be related to specific molecular, cellular or tissue functions and the biological components (i.e. genes) are involved in multiple functions.\\

@@

The current network represented by the adjacency matrix $A$, corresponds to a complete and weighted network of $8,928$ genes (nodes) and $39,850,128$ edges. For computational reasons, this network was transformed into an unweighted one $\hat{A}$, keeping only the connections above the cutoff value of $0.2$. The resulting unweighted network has a total of $5,810$ nodes and $16,875,145$ edges. After applying the HLC algorithm, a total of $4,131$ genes were distributed in $5,143$ overlapping modules of $3$ or more genes. Figure~\ref{fig:overlap} shows a histogram of the overlapping percentage of these genes, measured as the proportion of modules to which each gene belongs.\\

@@

\subsection{Modules association to phenotypic traits}

Our approach to identify the most relevant gene groups (modules) in the response to salt stress, consists in relating the co-expression network with phenotypic data from the samples. We use 3 phenotypic traits: shoot $K^+$ content, root biomass and shoot biomass. These were measured for the 92 genotypes studied, under  control and salt stress conditions. These data can be found in the supplementary information of~\cite{campbell2017allelic}. As shown in Figure \ref{fig:pdata}, there are significant differences in the values of these phenotypic traits between both stress and control conditions. This supports the idea that these variables represent tolerance-associated traits in rice under salt stress.\\

@@

%Módulos seleccionados + cantidad de genes por módulo

\subsection{Genes enrichment}
% genes seleccionados con expresión diferencial según análisis DEA. 
% gene ontology de esos genes

\section{Results}
% redes de los genes anotados

>>>>

\section*{Introduction}
Abiotic stresses are the key factors which negatively influence plant development and productivity. They are the main cause of extensive agricultural production losses worldwide. One of the most devastating abiotic stresses, causing reduction in the cultivable land, crop quality and productivity is soil salinity. It has been estimated that 20\% of total cultivated and 33\% of irrigated agricultural lands worldwide are already affected by high salinity. Due to the human activities and natural causes, salinized areas are gradually increasing every year and are expected to reach 50\% by the end of year 2050~\cite{shrivastava2015soil}. Salinity effects are the result of elaborated interactions among morphological, physiological, and biochemical processes. Those processes are regulated by multiple genes and determine the salt tolerance or susceptibility of the crop~\cite{reddy2017salt}. Thus, identifying this group of stress responsive genes may lead to crop improvement in salt tolerance, which is known as a complex quantitative trait. Find this target genes is a complex task, because the function of many genes are still not understood and many novel non-coding genes have been discovered. Particularly rice (\textit{Oryza sativa}), the major food source around the world, is highly sensitive to salt stress~\cite{chang2019morphological}. Therefore, identification of target genes in rice may allow biologist use them as a genetic resource to develop new cultivars with resistance to salinity.\\

We propose a methodology to identify stress responsive genes to salt conditions in rice. The methodology is based on Weighted Gene Co-expression Network Analysis (WGCNA). This is considered an effective and accurate bioinformatic method, using co-expression networks, that has been widely applied in identifying target genes for disease and cancer fields~\cite{tian2018identifying}. We follow the WGCNA workflow but with a new approach in the module detection step. Our modules are detected using the Hierarchical Link Clustering (HLC) technique~\cite{ahn2010link} that allows the recognition of overlapping communities, which may have more biological meaning given the overlapping regulatory domains of systems that generate co-expression~\cite{gaiteri2014beyond}. We conduct a systematic study with a large set of rice data using the proposed methodology. RNA-seq data was accessed through GEO database~\cite{GEOAcces90:online} (Accession number GSE98455), corresponding to $57845$ gene expression profiles of shoot tissues measured for both control and salt condition in $92$ accessions of the Rice Diversity Panel 1. As the analysis result, 6 modules are detected as relevant in the response to salt stress in rice: 3 modules of 3 genes each one associated with shoot K content, 2 modules of 3 genes associated with  shoot biomass, and 1 module of 4 genes associated with root biomass. These genes may act as potential targets for the improvement of salinity tolerance in rice cultivars. From those 19 genes, all but 3 genes (associated with $K$ content), were also identified as deferentially expressed ($|LFC| > 2$) for at least one of the 92 accessions, suggesting that those genes are strong candidates as stress responsive genes. Only 2 of the 16 diferentially expressed genes, both from the module related with shoot biomass, are named and have an associated protein product: Spermidine hydroxycinnamoyltransferase 2 (SHT2) and Lipoxygenase. In other words, further studies are needed to elucidate the detailed biological function of the remaining 14 genes that have not been named so far,  which may have a potential relevance in stress responsive mechanisms to salt conditions in rice.\\

With the development of high-throughput technologies, including microarrays and RNA sequencing (RNA-seq), genome-wide gene expression can be studied under different environmental stimuli (e.g. salt stress). Our methodology uses this kind of transcriptomic data measured for two different conditions (control and stress). After a process of normalization and filtering of the raw data, a differential expression profile of the genes is built calculating the log fold change (LFC) from control to stress condition. The LFC matrix will be the input for the co-expression network construction trough the WGCNA method. A similarity matrix is calculated using the absolute value of Pearson's correlation coefficient between pairs of genes. Then, the similarity matrix is forced to be a scale-free network, finding a beta exponent such that by raising each entry of the matrix to that value, the probability distribution follows a power law. Next, unlike WGCNA, the scale-free network is used to detect overlapping rather than non-overlapping communities, using the HLC technique. We also implement a LASSO regression~\cite{tibshirani1996regression} to select the most significant modules associated with rice phenotypical responses to salt stress. Finally, for the genes found, we look for previous evidence of important biological implications in tolerance to salt stress. That is, the genes deferentially expressed within the selected modules are enriched with gene ontology annotations from QuikGO database~\cite{binns2009quickgo} and their interaction networks reported in STRING database~\cite{szklarczyk2016string} are reviewed.\\

The proposed methodology is modular, since other module detection and selection techniques could be used, instead HLC and LASSO respectively. The advantage of using HLC as clustering method is its ability to detect overlapping modules, since biological components are involved in multiple functions and therefore biological communities tend to be highly overlapping ( ). On the other hand, LASSO is a regularized regression technique widely used in variable selection, thanks to its ability to obtain zero regression coefficients for the less relevant variables~\cite{desboulets2018review}. Additionally, LASSO is especially useful in problems where the number of variables is much larger than the number of samples, which is our case having more than 5000 modules (variables) and 92 accession (samples). The combinations of these techniques would allow finding target genes for future biological studies that evaluate their potential as genes that respond to salt stress in rice. Furthermore, this study can be extended to other stresses and even to other crops.\\

\section{Methodology}
\subsection{Data pre-processing} 
The RNA-seq data cannot be directly interpreted, therefore a normalization process has to be done to deal with the various biases that affect quantification results. The normalization technique used was DESeq2 \cite{love2014moderated}. From the normalized data, the two biological repetitions of each accession were averaged and genes exhibiting low variance (the ratio of upper quantile to lower quantile smaller than 1.5) or low expression (more than 80\% samples with values smaller than 10) were removed. At this point, control and salinity stress treatment data are separated into two matrices $C=[c_{ij}]_{n \times p}$ and $T=[t_{ij}]_{n \times p}$, respectively, where $c_{ij}$ and $t_{ij}$ represent normalized expression level of the gene $i$ in the accession $j$.\\

Changes in gene expression between control and stress conditions, are measured in terms of log ratios. Matrix $L=[e_{ij}]_{n \times p}$, known as the Log Fold Change matrix, is computed by setting $\ell_{ij}=\log_2 (t_{ij}/c_{ij})$.\\

@@

\subsection{Network construction}
The Log Fold Change matrix is used to construct the weighted co-expression network using the absolute value of the Pearson correlation as the similarity measure between genes. The matrix $S=[s_{ij}]_{n\times n}$ measures the level of concordance between gene expression profiles across experiments and is transformed into an adjacency matrix $A=[a_{ij}]_{n\times n}$ where $a_{ij} = (s_{ij})^\beta $ encodes the connection strength between each pair of genes. In other words, the elements of the adjacency matrix are the similarity values up to the power $\beta > 1$ so the degree distribution will fit a scale-free network. This kind of networks contain many nodes with very few connections and a small number of hubs with high connections. \\

@@

\subsection{Module detection}	
Once the network has been constructed, the next step is to study its structure and dynamics identifying communities also called modules. The idea is to cluster genes with similar expression change patterns. Membership in these modules may overlap in biological contexts, where modules may be related to specific molecular, cellular or tissue functions and the biological components (i.e. genes) are involved in multiple functions.\\

@@

The current network represented by the adjacency matrix $A$, corresponds to a complete and weighted network of $8928$ genes (nodes) and $39850128$ edges. For computational reasons, this network was transformed into an unweighted one $\hat{A}$, keeping only the connections above the cutoff value of $0.2$. The resulting unweighted network has a total of $5810$ nodes and $16875145$ edges. After applying the HLC algorithm, a total of $4131$ genes were distributed in $5143$ overlapping modules of $3$ or more genes. Figure~\ref{fig:overlap} shows a histogram of the overlapping percentage of these genes, measured as the proportion of modules to which each gene belongs.\\

@@

\subsection{Modules association to phenotypic traits}
Our approach to identify the most relevant gene groups (modules) in the response to salt stress, consists in relating the co-expression network with phenotypic data from the samples. We use 3 phenotypic traits: shoot $K^+$ content, root biomass and shoot biomass. These were measured for the 92 genotypes studied, under  control and salt stress conditions. These data can be found in the supplementary information of~\cite{campbell2017allelic}. As shown in Figure \ref{fig:pdata}, there are significant differences in the values of these phenotypic traits between both stress and control conditions. This supports the idea that these variables represent tolerance-associated traits in rice under salt stress.\\

@@

In our case study, 6 modules were detected as relevant in the response to salt stress in rice: 3 modules of 3 genes each one associated with shoot K content, 2 modules of 3 genes associated with  shoot biomass, and 1 module of 4 genes associated with root biomass.\\

\subsection{Genes enrichment}
% genes seleccionados con expresión diferencial según análisis DEA. 
% gene ontology de esos genes
% redes de interacción PP de los genes nombrados

### 2020-08-14 nicolas/PPIPL3/ReportV1.tex

%% LyX 2.3.5-1 created this file.  For more info, see http://www.lyx.org/.
%% Do not edit unless you really know what you are doing.
\documentclass[onecolumn,english]{article}
\usepackage[T1]{fontenc}
\usepackage[latin9]{inputenc}
\usepackage{float}
\usepackage{graphicx}

@@

\usepackage{babel}
\begin{document}
\title{Prediction of Protein-Protein Interactions on the Human and
Rice Interactome}
\author{Nicol\'as Antonio L\'opez Rozo}
\maketitle
\begin{abstract}
Networks science provides a set of computational tools to identify unmapped protein-protein interactions (PPI's). Barab\'asi et al. recently introduced a metric (based on network paths of length three) which 
proved to be an important feature in predicting links of the interactome network. In other words, paths of length three (L3) enable us to discover hidden link candidates for experimental validation. This paper extends the previous approach by Barab\'asi et al. using \texttt{XGBoost} module of Machine Learning (ML). We train the model using handcrafted features as well as features learned from embeddings. For the feature extraction from the network, \texttt{Node2Vec} is applied. Results show that while L3 feature remains important for predicting links, so are the features retrieved from the network structure using \texttt{Node2Vec}. The mentioned approaches are compared using the human and the rice interactomes.
\end{abstract}

\section{Introduction}

Proteins, as a fundamental constituents of any living being and their
``building blocks'' in terms of biological functionality, interact
among themselves and with other molecules according to their amino
acid sequences and specific three-dimensional structure and folding.
These interactions also depend on many physical and chemical properties
of each macro-molecule and on the cellular organelle where they operate,
such as pH and temperature. For this reason, experimental ways to
reproduce all of the necessary conditions to validate if an interaction
actually occurs within the cell are generally expensive.

@@

Previous result comes as no surprise when the biological basis of
protein interactions is considered: It is necessary that protein A
and protein B have complementary structures in order to interact,
and when classical length-2 paths are used, the predicted protein
interactions usually have the same structures, and not complementary
ones.

For the analyzed networks, it is also important to highlight their
robustness when compared to a random network with the same density:
if weighted spectral distribution (WSD) is used, then the used networks
are on average twice as robust as their random network versions.

\bibliographystyle{plain}
\bibliography{\string"refs\string"}

>>>>

%% LyX 2.3.5-1 created this file.  For more info, see http://www.lyx.org/.
%% Do not edit unless you really know what you are doing.
\documentclass[onecolumn,english]{article}
\usepackage[T1]{fontenc}
\usepackage[latin9]{inputenc}
\usepackage{float}
\usepackage{graphicx}
\usepackage{todonotes}

@@

\usepackage{babel}
\begin{document}
\title{Prediction of Protein-Protein Interactions on the Human and
Rice Interactome}
\author{Nicol\'as Antonio L\'opez Rozo}
\maketitle
\begin{abstract}
Previous Network-based efforts to predict unmapped protein-protein interactions (PPI's) suggest that proteins with multiple paths of length~3 (L3) are more likely to be connected. This paper extends this so-called L3 principle by taking into account feature extraction and using \texttt{XGBoost} techniques for prediction. In particular, we train the model using handcrafted features as well as features learned from embeddings using \texttt{Node2Vec}. Our main result shows that while L3 remains an important principle for predicting links, the approach is outperform by using embedded features. The mentioned approaches are compared using the human and the rice interactomes.

% need include a feature that compares the dimension of the embedding is useful for prediction with L3 to show that they independent measures - we do not want to be learning L3 through node2vec
\end{abstract}

\section{Introduction}
% need to add other relevant references 
Proteins, as a fundamental constituents of any living being and their
``building blocks'' in terms of biological functionality, interact
among themselves and with other molecules according to their amino
acid sequences and specific three-dimensional structure and folding.
These interactions also depend on many physical and chemical properties
of each macro-molecule and on the cellular organelle where they operate,
such as pH and temperature. For this reason, experimental ways to
reproduce all of the necessary conditions to validate if an interaction
actually occurs within the cell are generally expensive.

@@

Previous result comes as no surprise when the biological basis of
protein interactions is considered: It is necessary that protein A
and protein B have complementary structures in order to interact,
and when classical paths of length~2 are used, the predicted protein
interactions usually have the same structures, and not complementary
ones.

% please update the conclusions - WSD does not seem relevant 
For the analyzed networks, it is also important to highlight their
robustness when compared to a random network with the same density:
if weighted spectral distribution (WSD) is used, then the used networks
are on average twice as robust as their random network versions.

\bibliographystyle{plain}
\bibliography{refs.bib}
